%!TEX root = paper.tex
%
% PyClaw
%
% Lead currently:  David Ketcheson
%

\subsection{\pyclaw} \label{sec:pyclaw}

\pyclaw is a Python-based version of full-fledged Python solvers in which
higher-level parts of \clawpack have been reimplemented in Python and adapted to
an object-oriented framework.  \pyclaw's data structures and IO routines are
also used by the other \clawpack codes and VisClaw for post-processing, a set of
capabilities that was first implemented in later versions of 4.x including a
primitive version of the solver.  The new and fully implemented solver also
includes access to the high-order algorithms introduced in SharpClaw (i.e., WENO
reconstruction and Runge--Kutta integrators) and can be used on large
distributed-memory parallel machines for both the high-order algorithms and the
original Lax-Wendroff based sovlers.  Lower-level code (whatever gets executed
repeatedly and needs to be fast) from the earlier Fortran Classic and SharpClaw
codes is automatically wrapped at install time using \texttt{f2py}.

\subsubsection{Librarization and Extensibility}
Scientific software is easier to use, extend, and integrate with other tools
when it is designed as a library \cite{Brown:2015cj}.  \clawpack has always been
designed to be extensible, but \pyclaw takes this further in several ways.
First, it is distributed via a widely-used package management system,
\texttt{pip}. Second, the default installation process (``\texttt{pip install
clawpack}'') provides the user with a fully-compiled code and does not require
setting environment variables.  Like other \clawpack packages, \pyclaw provides
several ``hooks'' for users to plug in custom routines (for instance, to specify
boundary conditions). In \pyclaw, these routines -- including the Riemann solver
itself -- are selected at run-time, rather than at compile-time.  These routines
can be written directly in Python, or (if they are performance-critical) in a
compiled language (like Fortran or C) and wrapped with one of the many available
tools.  Problem setup (including things like initial conditions, algorithm
selection, and output specification) is also performed at run-time, which means
that researchers can bypass much of the slower code-compile-execute-post-process
cycle. It is intended that \pyclaw be easily usable within other packages
(without control of \texttt{main()}).  This is in contrast to \amrclaw, \geoclaw
and classic, which provides an end-to-end solution.

\subsubsection{Python classes for data on collections of structured grids}
\pyclaw includes Python classes for describing collections of structured grids
and data on them. These classes are also used by the other codes through
VisClaw, for post-processing.

A mesh in \clawpack always consists of a set of (possibly) mapped tensor-product
grids (interval, quadrilateral, or hexahedral).  In \pyclaw this geometry is
represented through the following objects:
\begin{itemize}
    \item A {\bf Dimension} is an interval divided into a fixed number of equally-sized subintervals.
    \item A {\bf Patch} is the tensor product of 1-3 Dimension objects, possibly with an associated mapping.
    \item A {\bf Domain} is a collection of any number of possibly overlapping Patch objects.
\end{itemize}
At present, \pyclaw solvers operate only on a single patch.  Multi-patch
capability is used to represent geometries from \amrclaw and \geoclaw runs, and
is planned for utilization in a future AMR-enabled version of \pyclaw.

The simulation data are represented by an additional set of objects:
\begin{itemize}
    \item A {\bf State} contains values of the solution vector $q$ and any
        auxiliary variables for each cell in a given Patch.
    \item A {\bf Solution} is a collection of States, corresponding to values of the solution
        on each Patch in a Domain.
\end{itemize}

\subsubsection{\pyclaw Solvers} A Solver is used to advance a Solution in time.
Two types of solvers are implemented: 
\begin{itemize}     
    \item {\bf Classic}:  the Lax-Wendroff-based solvers that are used in other
    Clawpack codes.     
    \item {\bf SharpClaw}: high-order solvers based on WENO reconstruction and
    Runge-Kutta or linear multistep time integration. 
\end{itemize} 
All \pyclaw Solvers can use
both fixed and variable step-size. In order to preserve stability a CFL-like
condition is being checked at each step of the computation, based on a user-
defined maximum CFL number. The step-size must be chosen in such a way that the
CFL number is always less than the maximum allowed  value, thus in the case of
variable step-size a rigorous step-size control is required. We distinguish the
Lax-Wendroff-based solvers and Runge-Kutta methods from linear multistep
methods, since the step-size strategy is different.

For the Lax-Wendroff and Runge-Kutta methods, first a choice for the step-size
is made according to the  maximum wave speed of the current solution  (and the
maximum wave speeds of stage solutions for Runge--Kutta methods). Then a new
solution is computed using this step-size and the CFL number is evaluated. If
the CFL number is less than the maximum allowed value then the step is accepted
and a new step-size is  chosen in the same manner. Otherwise, the wave speed at
the current step is large enough to cause instability or spurious oscillations
if  such a step-size is used, hence the step is rejected. An updated step-size
based on the increase of the maximum wave speed is calculated and the step is
repeated. This guarantees that a step is rejected at most once and in most cases
only the first step is rejected due to  large initial step-size. Currently
SharpClaw Runge--Kutta time integrators consist of Euler's method, second-,
third- and forth-order optimal SSP Runge--Kutta methods with two, three and ten
stages respectively. Also a general Runge-Kutta method with user-defined
coefficients can be used.

For linear multistep methods the step-size selection is more involved since a
variable step-size depends on the SSP coefficient of the method which in turn
depends on previous step-sizes and the method's coefficients  that are varied to
attain hight order.  All these intricate relations are analyzed in \todo{cite
SSPVLMM paper}, where second and third order SSP linear multistep methods with
variable step-size are developed. SharpClaw linear multistep methods consist of
second-order, three-step and third-order, four-step optimal  SSP linear
multistep methods. In both cases explicit formula for the step-size are used
based on the permissible SSP step-size that  guarantees the SSP property. SSP
Runge--Kutta are used for the starting values and the step-size for the those
steps is chosen as explained above. For the third-order linear multistep methods
additional {\em a posteriori} conditions must be applied. Basically, they imply
that the step-size remains finite and non-zero and that the forward Euler step-
size changes little over a single step. Whenever these conditions are violated
the step is rejected, a new step-size is derived and the solution is
recalculated. Again, the fix in the step-size is such that the step is rejected
at most once.

SharpClaw uses a traditional method-of-lines approach, in which
spatial operators are discretized first, to achieve high-order
resolution in space and time.  Spatial derivatives are discretized
using a high-order non-oscillatory reconstruction of the solution from
cell averages, which results in a system of ODEs that are subsequently
solved using the time-stepping methods described above.  The spatial
reconstructions are obtained using the weighted essentially
non-oscillatory (WENO) reconstruction algorithm which suppresses
spurious oscillations near discontinuities.  The WENO routines in
SharpClaw were generated by PyWENO, which is a standalone package for
building custom WENO codes. \cite{ketcheson2012pyclaw} 

\todo{should a
  brief description of PyWENO appear here, or just a citation?}