%!TEX root = paper.tex
%
% Riemann
%
% Lead currently:  David
%

\subsection{Riemann: A Community-Driven Collection of Approximate Riemann Solvers}
The methods implemented in Clawpack, and all modern Godunov-type methods for
hyperbolic PDEs, are based on the solution of Riemann problems.  A Riemann
problem is the Cauchy initial value problem posed by a first-order system of
hyperbolic PDEs and a piecewise-constant initial state with a single
discontinuity.  Riemann problems are central to hyperbolic PDEs because
discontinuities arise spontaneously in their solution.  The main theoretical
and numerical difficulties of hyperbolic problems involve the prescription of
physically correct weak solutions -- in other words, understanding the behavior
of the solution at discontinuities.  The Riemann solver is an algorithm that
encodes the specifics of the hyperbolic system to be solved, and it is the only
routine (other than problem-specific setup) that needs to be changed in order
to apply the code to different hyperbolic systems.  In some cases, the Riemann
solver may also be designed to enforce physical properties like positivity
(e.g., for the water depth in GeoClaw) or to account for forces (like that
of gravity) that may be balanced by flux terms.

For nonlinear systems, the exact solution of the Riemann problem is computationally
costly and may involve both discontinuities (shocks and contact waves) and
rarefactions.  It is almost always preferable to employ inexact Riemann solvers
that approximate the solution using discontinuities only.  Only the latter type
of Riemann solvers may be used in Clawpack.  For some systems, like the Euler
equations of compressible flow, a range of approximate solvers (giving 
more or less accurate results) has been developed in the literature.

One of the main commonalities across all packages in the Clawpack suite is the
use of a standard interface for Riemann solver routines.  This ensures that new
solvers or solver improvements developed for one package can immediately
be used by all packages.  To further facilitate this sharing and to avoid 
duplication, Riemann solvers are (with rare exceptions) not maintained under
the other packages but are collected in a single repository named Riemann.
Users who develop new solvers are strongly encouraged to submit them to the
Riemann repository.

In the Fortran-based packages (Classic, AMRClaw, and GeoClaw) the Riemann
solver is selected at compile-time by modifying a problem-specific Makefile.
In PyClaw, the Riemann solver to be used is selected at run-time.  This is
made possible by compiling all of the Riemann solvers (when PyClaw is installed)
and generating Python wrappers with f2py.  For PyClaw, Riemann also provides
useful metadata (such as the number of equations, the number of waves, and
the names of the conserved quantities) for each solver.

Whereas most existing codes for hyperbolic PDEs use Riemann solvers to
compute fluxes, Clawpack Riemann solvers instead compute the waves 
(or discontinuities) that make up the approximate Riemann solution.
Clawpack is also unusual in that it optionally makes use of transverse
Riemann solvers, responsible for computing transport between cells that
touch only along an edge or a corner.
